Stochastic Approximation Monte Carlo with a Dynamic Update Factor

### 2019-08-08 papers/histogram/sad-referee-response.md

Response to first referee

The first referee acknowledges that our paper is well written and that the field
is an important one, but does not believe that our work is a *'substantial
contribution*'' to the field.  The referee has three major criticisms of the
value of our work (aside from issues with the introduction section).

Our method is based on SAMC (as we clearly state in the paper) and is a substantial
improvement on this method (convergence is of 1 - 2 orders of magnitude superior
as can be seen from Fig. 3 and Fig. 4).  In addition, we also provide a detailed
comparison with 4 different methods on the square well system.

Firstly, the referee criticizes our paper for testing the method on a single
physical system (the square well fluid).  The referee gives as an argument an
example of a paper which tests a method on three different physical systems.  We
give as counter examples *references that we cite* which each use a single
physical system as the test case.  We agree that our method should be tested on
additional different physical systems but dispute that this is necessary for the
first paper introducing an improved method.

For instance, our citations
[10,15,18,19,22] are all papers (the first three published in PRE **?**) which
examine the convergence properties of flat histogram methods using a single
physical system.

**Avoiding boundary effects in Wang-Landau sampling** *(accepted by PRE)*
This paper with DP Landau as an author considers a single system (the 2D Ising
model). The novelty of the paper is providing a scheme to avoid boundary effects
in WL. [10]

**Fast Calculation of the Density of States of a Fluid by Monte Carlo Simulations**
*(accepted by PRL)*
This paper compares two new methods with WL on the truncated Lennard-Jones system. [11]

**Optimizing the ensemble for equilibration in broad-histogram Monte Carlo simulations**
*(accepted by PRE)*
This paper presents an adaptive algorithm and compares with WL on 2D Ising systems. [12]

**Analysis of the convergence of the 1/t and Wang-Landau algorithms in the
calculation of multidimensional integrals** *(accepted by PRE)*
This paper compares the performance of two methods (WL and 1/t-WL) to calculate
different integrals (not a real physical system because the exact answer is
known). [15]

**Optimal modification factor and convergence of the Wang-Landau algorithm**
*(accepted by PRE)*
This paper examines the optimal modification factor for the WL algorithm and
tests on a single system (the 2D Ising model). [18]

**Convergence of Stochastic Approximation Monte Carlo and modified Wang–Landau
algorithms: Tests for the Ising model**
*(accepted by Computer Physics Communications)*
This paper compares SAMC, WL, and 1/t-WL on a single system (the 2D Ising model). [19]

**Stochastic approximation Monte Carlo and Wang–Landau Monte Carlo applied to a
continuum polymer model** *(accepted by Computer Physics Communications)*
SAMC and WL are compared for a single continuum model with a square well type
potential. [22]

**Direct calculation of liquid–vapor phase equilibria from transition
matrix Monte Carlo simulation** *J. Chem. Phys.*
This paper introduces a new method (adaptation of Fitzgerald et al.) TMMC and tests
on Lennard-Jones fluid. [34]

**Performance Limitations of Flat-Histogram Methods**
*(accepted by PRL)*
This paper examines optimal scaling of flat histogram methods on 2D Ising systems. [35]

**Flat Histogram Methods for Quantum Systems: Algorithms to Overcome Tunneling Problems
and Calculate the Free Energy**
*(accepted by PRL)*
This paper examines the generalization of WL algorithm on a spin-1/2 Heisenberg model. [36]

**Adaptive integration method for Monte Carlo simulations**
*(accepted by PRE)*
This paper introduces an adaptive sampling method and tests on Lennard-Jones. [37]

**Dynamically optimized Wang-Landau sampling with adaptive trial moves and
modification factors**
*(accepted by PRE)*
This paper introduces dynamically optimized WL sampling and tests on XY-Model. [38]

### 2019-08-08 papers/histogram/sad.tex
added more examples of papers that test on only a single system.

\section{Introduction}
Over the past several decades a number of flat histogram Monte Carlo simulation
algorithms have been developed which calculate the thermodynamic properties of
various systems for over a range of temperatures.  This development began with
the original histogram method, which used a single canonical Monte Carlo
simulation to predict properties for nearby
temperatures~\cite{ferrenberg1988new}.  For large systems this approach is
limited to a narrow temperature range because a single canonical simulation
explores only a small range of energies. Berg and Neuhaus developed multicanonical
methods that introduced a weight function to enable flat histogram sampling
which improved the exploration of configuration space and allows the system
to
overcome free energy barriers~\cite{berg1991multicanonical, berg1992multicanonical}.
These works led to an explosion in the interest of
``flat'' (or
``broad'') histogram methods~\cite{penna1996broad, penna1998broad,
swendsen1999transition, wang2001determining, wang2001efficient, landau2004new, schulz2003avoiding, yan2003fast,
trebst2004optimizing,
belardinelli2007wang, belardinelli2007fast,
belardinelli2008analysis, belardinelli2014intrinsic, singh2012density, zhou2008optimal,
schneider2017convergence, liang2006theory, liang2007stochastic,
kim2006statistical,
kim2007statistical, kim2009replica, junghans2014molecular}, which explore a wider range of energies.  Another
benefit, in contrast with low-temperature canonical Monte Carlo, is that these
approaches cannot be trapped in a local energy minimum.

@@

Liang began to consider that WL could be considered a special case of the
Stochastic Approximation approach, whose convergence could be mathematically
proven~\cite{liang2006theory}. In 2007, Liang, Liu, and Carrol \emph{et al.}
developed Stochastic Approximation Monte Carlo
(SAMC)~\cite{liang2007stochastic}, and proved its convergence, although the
method still has a system specific user-defined parameter which must be tuned
when applying this algorithm to a new system.  In
contrast to the WL-based methods, the SAMC method does not require users to
determine the energy range of interest \emph{a priori}.

>>>>

\section{Introduction}
Over the past several decades a number of flat histogram Monte Carlo simulation
algorithms have been developed which calculate the thermodynamic properties of
various systems for over a range of temperatures.  This development began with
the original histogram method, which used a single canonical Monte Carlo
simulation to predict properties for nearby
temperatures~\cite{ferrenberg1988new}.  For large systems this approach is
limited to a narrow temperature range because a single canonical simulation
explores only a small range of energies. Berg and Neuhaus developed multicanonical
methods that introduced a weight function to enable flat histogram sampling
which improved the exploration of configuration space and allows the system
to
overcome free energy barriers~\cite{berg1991multicanonical, berg1992multicanonical}.
These works led to an explosion in the interest of
``flat'' (or
``broad'') histogram methods~\cite{penna1996broad, penna1998broad,
swendsen1999transition, wang2001determining, wang2001efficient, landau2004new, schulz2003avoiding, yan2003fast,
trebst2004optimizing,
belardinelli2007wang, belardinelli2007fast,
belardinelli2008analysis, belardinelli2014intrinsic, singh2012density, zhou2008optimal,
schneider2017convergence, liang2006theory, liang2007stochastic, werlich2015stochastic,
kim2006statistical,
kim2007statistical, kim2009replica, junghans2014molecular}, which explore a wider range of energies.  Another
benefit, in contrast with low-temperature canonical Monte Carlo, is that these
approaches cannot be trapped in a local energy minimum.

@@

Liang began to consider that WL could be considered a special case of the
Stochastic Approximation approach, whose convergence could be mathematically
proven~\cite{liang2006theory}. In 2007, Liang, Liu, and Carrol \emph{et al.}
developed Stochastic Approximation Monte Carlo
(SAMC)~\cite{liang2007stochastic}, and proved its convergence, although the
method still has a system specific user-defined parameter which must be tuned
when applying this algorithm to a new system.  In
contrast to the WL-based methods, the SAMC method does not require users to
determine the energy range of interest \emph{a priori}.
Werlich \emph{et al.} proposed the introduction of an additional tuning parameter
into SAMC~\cite{werlich2015stochastic}.
